Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-18 BN_discretizer_paper.tex
more notation changes

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
  %!PS-Adobe-3.0 EPSF-3.0
  %%BoundingBox: 19 19 221 221
  %%CreationDate: Mon Sep 29 1997
  %%Creator: programmed by hand (JK)
  %%EndComments
  gsave
  newpath
    20 20 moveto
    20 220 lineto
    220 220 lineto
    220 20 lineto
  closepath
  2 setlinewidth
  gsave
    .4 setgray fill
  grestore
  stroke
  grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepgfplotslibrary{groupplots}
\usepackage{tikz}
\usepackage{lscape}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{booktabs}
% \usepackage[capitalize]{cleveref}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\pgfplotsset{
    jitter/.style={
        x filter/.code={\pgfmathparse{\pgfmathresult+rnd*#1}}
    },
    jitter/.default=0.1
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CUSTOM COMMANDS

@@

\DeclareMathOperator{\Pa}{pa}

@@

\institute{F. Author \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{fauthor@example.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           S. Author \at
              second address
}

@@

% \begin{table}[H]
%   \centering
%   \begin{tabular}{@{}ccl@{}}
%     \toprule
%     Current & Mykel & Description \\
%     \midrule
%     $G$ & $G$ & BN graph structure \\
%     $P$ & $P$ & discrete probability \\
%     $p$ & $p$ & probability density \\
%     $D$ & $D$ & available data \\
%     $|D|$ & $n$ & number of variables in the BN \\
%     $D_{\bm{Y}}$ & & subset of data associated with $\bm{Y}$ \\
%     $X$ & $X_{1:n}$ & variables in BN (discrete) \\
%     $x_{1:n}$ & $x_{1:n}$ & a specific instance of vars in BN \\
%     $x'_{1:n}$ &  & unique instances of var in BN \\
%         & $r_i$ & number of instantiations of $i$th var \\
%     $J_p$ & $q_i$ & number of instantiations of parents of $i$th var \\
%         &       & $ = 1$ if no parents \\
%     $\brock{\Pi_{X_i}}_j$ & $\pi_{ij}$ & $j$th instantiation of $i$th var \\
%         & $\theta_{ijk}$ & probability of $i$th var $= k$ given $j$th parental inst. \\
%     $\Theta$ & $\bm{\theta}$ & set of all (non-independent) probability parameters \\
%     $n_{i,j}^{(k)}$ & $m_{ijk}$ & observation counts \\
%     $n_{i,j,l}^{(k)}$ & $m_{ijk}$ & observation counts with $l$th instance of spouse\\
%     $\alpha_{ijk}$ & $\alpha_{ijk}$ & Dirichlet prior counts \\
%     $N_{ij}$ & $\alpha_{ij0}$ & sum over Dirichlet prior counts for $k$ \\
%         & $m_{ij0}$ & sum over observations counts for $k$ \\
%         & $\Gamma$ & the Gamma function \\
%     $N$ &          & is the number of samples in dataset \\
%     $N_i$ &          & counts for $i$th instance of a specific var \\
%     $\Pi_X$ & $\text{pa}_{x_i}$ & parents of a variable or instance \\
%     $n_C$ & & number of children of a variable \\
%     $H(~)$ & & a function \\
%     $I(~,~)$ & & mutual information \\
%     $X_i$ & & $i$th variable in BN, may be cont or disc \\
%     $X_i^*$ & & discrete version of cont var \\
%     $E$ & & set of edges in BN \\
%     $\text{Pa}\paren{X_i}$, $n_p$ & & number of parents of $X_i$ \\
%     $f\paren{X_i, \Pi_{X_i}}$ & $f(G)$ & score function for K2 \\
%     $M_C$ & & discretization policy for $C$ \\
%     $C$ & & a continuous variable \\
%     $t_i$ & & the $i$th bin edge for a specific var \\
%     $d_i$ & & allowable discretization positions (between unique vals) \\
%     $k$ & & number of intervals for $M_C$ \\
%     $C_{1:N}$ & & sorted samples for $C$ \\
%     $n_i$ & & $n_1 = s_1$ and $n_i = s_{i} - s_{i-1}$ \\
%     $s_i$ & & $\sum_{j=1}^k n_j$ and $s_k = N$, $s_0 = 0$ \\
%     $C_i$ & & $i$th child of a specific var \\
%     $\boldsymbol{S_i}$ & & set of spouses associated with $i$th child of specific var \\
%     $L$ & & largest cardinality among discrete variables in BN \\
%     $I_q$ & & an interval from $u \rightarrow v$ \\
%     $v'_\text{max}$ & & largest cardinality of variables that directly connect to $X$ \\
%     $S[]$ & &  optimal objective value of subproblems with instances from $1$ to $u$ \\
%     \bottomrule
%   \end{tabular}
% \end{table}

@@

MODL discretization \citep{Boulle_2006} is a Bayesian method for discretizing a continuous feature according to a class variable.
Bayesian methods maximize the probability of the model given the data, $P(\textit{Model} \mid \textit{Data})$.
An application of Bayes' rule and constant $P(\textit{Data})$ shows that this is equivalent to maximizing:

@@

A Bayesian network $B$ is defined by a pair $(G,\bm{\theta})$, where $G$ is a directed acyclic graph whose $n$ nodes $X_{1:n}$ are random variables.
The edges in a Bayesian network represent probabilistic dependencies among nodes and encode the Markov property: each node $X_i$ is independent of its non-descendants given its parents $\Pa_{X_i}$ in $G$.
For a Bayesian network over discrete and discretized variables, let $r_i$ represent the number of instantiations of $X_i$, and let $q_i$ represent the number of instantiations of the parents of $X_i$.
If $X_i$ has no parents, then $q_i = 1$.
The $j$th instantiation of the parents of $X_i$ is denoted $\bm{\pi}_{ij}$.

@@

@@

\todo{CONTINUE HERE}
A discretization policy $M_C = \anglebrackets{t_1 < t_2 < \ldots < t_{k-1}}$ for a continuous variable $C$ is a mapping from $\mathbb{R}$ to $\brock{1,2,3,\ldots,k}$ such that
\begin{equation}
  M_C (x)=\begin{cases}
    1, & \text{if $x<t_1$}.\\
    i, & \text{if $t_{i-1} \leq x < t_i$}\text{.}\\
    k, & \text{if $t_{k-1} \leq x$}
  \end{cases}
\end{equation}

\noindent
The policy discretizes the continuous variable $C$ into $k$ intervals.
The discretization edges $t_{1:k-1}$ are restricted to the midpoints between two sorted instances of $C$.
An integer representation of $M_C$ can be obtained by sorting the samples of $C$ in an increasing order, $\curly{c_1,c_2,\ldots,c_N}$, and if $t_i = (c_{s_{i}} + c_{s_{i}+1})/2$ for $i=1,2,\ldots,k$, then $M_C = \anglebrackets{t_1,t_2, \ldots, t_{k-1}} \equiv \brock{n_1,n_2,..n_k}$, where $n_1 = s_1$ and $n_i = s_{i} - s_{i-1}$.

@@

This section covers discretization of a single continuous variable $X$ in a Bayesian network where all other variables are discrete.
Let $n_p$ and $n_c$ respectively refer to the number of parents and number of children of $X$.
Each child $C_i$ of $X$ may possess other parents, referred to as spouses of $X$.
The set of spouses of $X$ is denoted $\boldsymbol{S_i}$.

@@

For example, in Figure 1., the corresponding $P(D \mid M)$ is factorized according to
\begin{equation}
P\paren{D_{ P_1,P_2,P_3 } \mid M} \cdot P\paren{ D_{ C_1 } \mid M,D_{\boldsymbol{S_1}}} \cdot P\paren{D_{C_2} \mid M,D_{ S_2  }}\text{.}
\end{equation}

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph1}
    \end{tabular}
  \caption{Factorization of $P(D \mid M)$}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
  %!PS-Adobe-3.0 EPSF-3.0
  %%BoundingBox: 19 19 221 221
  %%CreationDate: Mon Sep 29 1997
  %%Creator: programmed by hand (JK)
  %%EndComments
  gsave
  newpath
    20 20 moveto
    20 220 lineto
    220 220 lineto
    220 20 lineto
  closepath
  2 setlinewidth
  gsave
    .4 setgray fill
  grestore
  stroke
  grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepgfplotslibrary{groupplots}
\usepackage{tikz}
\usepackage{lscape}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{boondox-cal} % for lowercase caligraphic
% \usepackage[capitalize]{cleveref}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\pgfplotsset{
    jitter/.style={
        x filter/.code={\pgfmathparse{\pgfmathresult+rnd*#1}}
    },
    jitter/.default=0.1
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CUSTOM COMMANDS

@@

\DeclareMathOperator{\Pa}{pa} % parents of a var in a BN
\newcommand{\cX}{\mathcal{X}} % continuous X
\newcommand{\cx}{\mathcal{x}} % instance of x
\newcommand{\cu}{\mathcal{u}} % unique sorted instances of x

@@

\institute{\todo{Yi-Chun Chen \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{fauthor@example.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           S. Author \at
              second address}
}

@@

\begin{table}[H]
  \centering
  \begin{tabular}{@{}ccl@{}}
    \toprule
    Current & Mykel & Description \\
    \midrule
    % $G$ & $G$ & BN graph structure \\
    % $P$ & $P$ & discrete probability \\
    % $p$ & $p$ & probability density \\
    % $D$ & $D$ & available data \\
    $|D|$ & $n$ & number of variables in the BN \\
    $D_{\bm{Y}}$ & & subset of data associated with $\bm{Y}$ \\
    $X$ & $X_{1:n}$ & variables in BN (discrete) \\
    $x_{1:n}$ & $x_{1:n}$ & a specific instance of vars in BN \\
    $x'_{1:n}$ &  & unique instances of var in BN \\
        & $r_i$ & number of instantiations of $i$th var \\
    $J_p$ & $q_i$ & number of instantiations of parents of $i$th var \\
        &       & $ = 1$ if no parents \\
    $\brock{\Pi_{X_i}}_j$ & $\pi_{ij}$ & $j$th instantiation of $i$th var \\
        & $\theta_{ijk}$ & probability of $i$th var $= k$ given $j$th parental inst. \\
    $\Theta$ & $\bm{\theta}$ & set of all (non-independent) probability parameters \\
    $n_{i,j}^{(k)}$ & $m_{ijk}$ & observation counts \\
    $n_{i,j,l}^{(k)}$ & $m_{ijk}$ & observation counts with $l$th instance of spouse\\
    $\alpha_{ijk}$ & $\alpha_{ijk}$ & Dirichlet prior counts \\
    $N_{ij}$ & $\alpha_{ij0}$ & sum over Dirichlet prior counts for $k$ \\
        & $m_{ij0}$ & sum over observations counts for $k$ \\
        & $\Gamma$ & the Gamma function \\
    $N$ &          & is the number of samples in dataset \\
    $N_i$ &          & counts for $i$th instance of a specific var \\
    $\Pi_X$ & $\text{pa}_{x_i}$ & parents of a variable or instance \\
    $n_C$ & & number of children of a variable \\
    $H(~)$ & & a function \\
    $I(~,~)$ & & mutual information \\
    $X_i$ & & $i$th variable in BN, may be cont or disc \\
    $X_i^*$ & & discrete version of cont var \\
    $E$ & & set of edges in BN \\
    $\text{Pa}\paren{X_i}$, $n_p$ & & number of parents of $X_i$ \\
    $f\paren{X_i, \Pi_{X_i}}$ & $f(G)$ & score function for K2 \\
    $M_C$ & & discretization policy for $C$ \\
    $C$ & & a continuous variable \\
    $t_i$ & & the $i$th bin edge for a specific var \\
    $d_i$ & & allowable discretization positions (between unique vals) \\
    $k$ & & number of intervals for $M_C$ \\
    $C_{1:N}$ & & sorted samples for $C$ \\
    $n_i$ & & $n_1 = s_1$ and $n_i = s_{i} - s_{i-1}$ \\
    $s_i$ & & $\sum_{j=1}^k n_j$ and $s_k = N$, $s_0 = 0$ \\
    $C_i$ & & $i$th child of a specific var \\
    $\boldsymbol{S_i}$ & & set of spouses associated with $i$th child of specific var \\
    $L$ & & largest cardinality among discrete variables in BN \\
    $I_q$ & & an interval from $u \rightarrow v$ \\
    $v'_\text{max}$ & & largest cardinality of variables that directly connect to $X$ \\
    $S[]$ & &  optimal objective value of subproblems with instances from $1$ to $u$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

MODL discretization \citep{Boulle_2006} is a Bayesian method for discretizing a continuous feature according to a class variable.
Bayesian methods maximize the probability of the model given the data, $P(\textit{Model} \mid \textit{Data})$.
An application of Bayes' rule and constant $P(\textit{Data})$ shows that this is equivalent to maximizing

@@

A Bayesian network $B$ is defined by a pair $(G,\bm{\theta})$, where $G$ is a directed acyclic graph whose $n$ nodes $X_{1:n}$ are random variables.
The edges in a Bayesian network represent probabilistic dependencies among nodes and encode the Markov property: each node $X_i$ is independent of its non-descendants given its parents $\Pa_{X_i}$ in $G$.
For a Bayesian network over discrete and discretized variables, $r_i$ represents the number of instantiations of $X_i$, and $q_i$ represents the number of instantiations of the parents of $X_i$.
If $X_i$ has no parents, then $q_i = 1$.
The $j$th instantiation of the parents of $X_i$ is denoted $\bm{\pi}_{ij}$.

@@

\todo{
Introduce P, C, and S for parents
}

@@

Let $\cX$ be a continuous variable and let $\cx$ be a specific instance of $\cX$.
A discretization policy $\Lambda_{\cX} = \curly{e_1 < e_2 < \ldots < e_{k-1}}$ for $\cX$ is a mapping from $\mathbb{R}$ to $\curly{1,2,3,\ldots,k}$ such that

\begin{equation}
  \Lambda_{\cX}(\cx) = \begin{cases}
    1, & \text{if $x<e_1$}\\
    i, & \text{if $e_{i-1} \leq x < e_i$}\\
    k, & \text{otherwise.}
  \end{cases}
\end{equation}

\noindent
The discretization policy discretizes $\cX$ into $k$ intervals.
Let the samples of $\cX$ in a given dataset $D$ be sorted in ascending order, ${\cx_{1:n} = \curly{\cx_1 \leq \cx_2 \leq \ldots \leq \cx_n}}$, and let the unique values be ${\cu_{1:n} = \curly{\cu_1 \leq \cu_2 \leq \ldots \leq \cu_n}}$.
The index of the last occurence of $\cu_i$ in $\cx$ is denoted $s_i$.

The discretization edges $e_{1:k-1}$ mark the boundaries between discretization intervals.
In this work they are restricted to the midpoints between unique ascending instances of $\cX$.
Thus, each edge $e_i = \paren{\cu_{s_{j+1}} - \cu_{s_j}}/2$ for some $s_j$.
An integer representation of $\Lambda_{\cX}$ can thus be written $\Lambda_{\cX} = \curly{\lambda_1 < \lambda_2  < \ldots < \lambda_k}$, where $\lambda_1 = s_1$ and $\lambda_i \in s_{1:|\cu|}$.

@@

This section covers the discretization of a single continuous variable $\cX$ in a Bayesian network where all other variables are discrete.
Let $n_p$ and $n_c$ respectively refer to the number of parents and number of children of $X$.
Each child $C_i$ of $X$ may possess other parents, referred to as spouses of $X$.
The set of spouses of $X$ is denoted $\boldsymbol{S_i}$.

@@

For example, in Figure~\ref{fig:example_factorization}, the corresponding $P(D \mid M)$ is factorized according to
\begin{equation}
P\paren{D_{ P_1,P_2,P_3 } \mid M} \cdot P\paren{ D_{ C_1 } \mid M,D_{\boldsymbol{S_1}}} \cdot P\paren{D_{C_2} \mid M,D_{ S_2  }}\text{.}
\end{equation}

\begin{figure}[ht]
  \label{fig:example_factorization}
  \begin{tabular}{cc}
    \input{graph1}
  \end{tabular}
  \caption{Factorization of $P(D \mid M)$}
\end{figure}
